---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# skater

<!-- badges: start -->
[![R-CMD-check](https://github.com/signaturescience/skater/workflows/R-CMD-check/badge.svg)](https://github.com/signaturescience/skater/actions)
<!-- badges: end -->

**S**NP-based **K**inship **A**nalysis **T**esting and **E**valuation: miscellaneous **R** utilties.

## Installation

Install from GitHub:

``` r
remotes::install_github("signaturescience/skater", auth_token = github_pat(quiet))
```

A GitHub [personal access token](https://github.com/settings/tokens) must be in the environment variable `GITHUB_PAT`. 

## Examples

```{r}
library(skater)
```

### Pedigrees and PLINK .fam files

Read a PLINK-formatted .fam file:

```{r}
famfile <- system.file("extdata", "3gens.fam", package="skater", mustWork=TRUE)
fam <- read_fam(famfile)
fam
```

Convert each family into a pedigree object (see the [kinship2 vignette](https://cran.r-project.org/web/packages/kinship2/vignettes/pedigree.html)):

```{r}
peds <- fam2ped(fam)
peds
```

In the example above, the resulting tibble is nested by family ID. The `data` column contains the individual family information, while the `ped` column contains the pedigree object for that family. You can unnest any particular family:

```{r}
peds %>% 
  dplyr::filter(fid=="testped1") %>% 
  tidyr::unnest(cols=data)
```

You can also look at a single pedigree:

```{r}
peds$ped[[1]]
```

Or plot that pedigree:

```{r plotped}
plot(peds$ped[[1]], mar=c(1,4,1,4))
```

The `plot_pedigree` function in the skater package will walk over a list of pedigree objects, writing a multi-page PDF to file, with each page containing a pedigree from each of the families:

```{r, eval=FALSE}
plot_pedigree(peds$ped, file="3gens.ped.pdf")
```

The `ped2kinpair()` function takes a pedigree object and produces a pairwise list of relationships with their expected kinship coefficient. 

Run on a single family:

```{r}
ped2kinpair(peds$ped[[1]])
```

Map over all families:

```{r}
kinpairs <- 
  peds %>% 
  dplyr::mutate(pairs=purrr::map(ped, ped2kinpair)) %>% 
  dplyr::select(fid, pairs) %>% 
  tidyr::unnest(cols=pairs)
kinpairs
```

Note that this maps `ped2kinpair` over all `ped` objects in the input tibble, and that relationships are not shown for between-family relationships (which should all be zero).

### Degree inference

The `dibble()` function creates a **d**egree **i**nference tibble, with degrees up to the specified `max_degree` (default=3), expected kinship coefficient, and lower (`l`) and upper (`u`) inference ranges (see [Manichaikul 2010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3025716/)). Degree 0 corresponds to self / identity / MZ twins, with an expected kinship coefficient of 0.5, with inference range >=0.354. Anything beyond the maximum degree resolution is considered unrelated (degree `NA`), with expected kinship coefficient of 0.

```{r}
dibble()
```

The degree inference `max_degree` default is 3. Change this argument to allow more granular degree inference ranges:

```{r}
dibble(max_degree = 5)
```

Note that the distance between relationship degrees becomes smaller as the relationship degree becomes more distant. The `dibble()` function will throw a warning with `max_degree` >=10, and will stop with an error at >=12.

The `kin2degree()` function infers the relationship degree given a kinship coefficient and a `max_degree` up to which anything more distant is treated as unrelated. Example first degree relative:

```{r}
kin2degree(.25, max_degree=3)
```

Example 4th degree relative, but using the default max_degree resolution of 3:

```{r}
kin2degree(.0312, max_degree=3)
```

Example 4th degree relative, but increasing the degree resolution:

```{r}
kin2degree(.0312, max_degree=5)
```

The `kin2degree()` function is vectorized over values of `k`, so can be used inside of a `mutate` on a tibble of kinship coefficients:

```{r}
# Get two pairs from each type of relationship we have in kinpairs:
kinpairs_subset <- 
  kinpairs %>% 
  dplyr::group_by(k) %>% 
  dplyr::slice(1:2)
kinpairs_subset

# Infer degree out to third degree relatives:
kinpairs_subset %>% 
  dplyr::mutate(degree=kin2degree(k, max_degree=3))
```

